---
title: "Republican Voteshare in the US Presidential Election"
author: "Raju Adhikari"
date: "March 16, 2017"
output: pdf_document
bibliography: sources.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r packages, echo=FALSE}

setwd("C:/RajuPC/MPP Final Thesis/MPP_Thesis_Adhikari")

```
#Literature Review:
The voteshare of an incumbment presidential candidate in a state is regressed against the general economic conditions of the state, which can be measured by the change in a state's unemployment rate and the percentage changes in real per capita personal income [@Abrams1980]. Previous vote share for the incumbent party is also introduced as an independent variable [@Abrams1980]. He claims that voters hold presidents accountable for changes in state-level economic conditions therefore cutting certain programs in strongly dominated states and moving them to closely contested states may improve the chances of relection for the incumbent [@Abrams1980]. He also claims that abolishing the electoral college and permitting popular votes to determine presidential outcomes can outweigh the efforts of such state-level redistributive policy [@Abrams1980]. The fact that Donald Trump's victory was highly dependent on the electoral college, since he lost the popular vote by almost XXXXXXXXX votes, and driven by a few contested states in the Rust Belt, show with even more significance the effect of regional economic conditions on vote share. 
_can this model be applied at a county level, since it is simply a smaller geographical setting and more representative since it is closer to the individual level_
_a dummy for Southern States was taken_ [@Abrams1980].

Lau and Redlawsk place "correct" voting as a middle ground between individual choice and social choice and therefore define it as the individual choice made under conditions of full information [@Lau1997].

Curry [@Curry2012] suggests that a lagged dependent variable would not be appropriate in this case since the relationship between the economy and the vote for president is not theorized as dynamic. He writes, "the data points are four years apart so it is unlikely that the performance of the economy at time t would be strong enough to significantly influence the vote for president four years in the future at time t + 1 or eight years in the future at time t + 2" [@Curry2012].



In economic voting, incumbents are given utmost importance because people often relate economic performance with the incumbent party or candidate. Hence, if the economy is doing well, it helps the incumbent party or candidate and if it is doing bad it hurts them.
_it would be interesting to see how people assign blame for their economic conditions on the President even if their regional governor or majority representation in the house may be from the non-incumbent's party._ 
Moreover, Curry [@Curry2012] writes that economic influence on the vote for president has been continuous and is not necessarily dependent on federal government control over the economy.


Three main inferences made by Bartels[@Bartels2005] are:
- The white working class has not abandoned the Democratic Party.
- The white working class has not become more conservative.
- Working class moral values do not trump economics.

Bartels [@Bartels2005] uses family income levels to categorize voters in terms of their economic status. He places families with incomes in the bottom third of the income distribution under "low income" or "working class" thus differentiating them from the middle and upper class families. 
_how do we operationalize this is terms of county-level economic data._

Bartels [@Bartels2005] illustrates the Democratic share of the two-party presidential vote among white voters in the bottom third of the income distribution and in the top third of the income distribution. He attributes the loss of support from white voters to Democratic candidates to the middle and upper income groups, while claiming that support for Democratic candidates has increased among low-income white voters. 
_see what happens when we see the actual county-level economic data on family income alongside Democratic or Republican voteshare_ 
_see what happens when we see the actual county-level economic data on family income, in areas with majority whites vs majority non-white, and see what happens_
_see what happens when we see the actual county-level economic data on family income, in areas with majority whites, in rural vs urban counties, and see what happens_

While survey data is useful in forecasting, the actual data we use can give a better picture of voter behavior since it captures the reality of after the election has taken place. However, based on the actual data, we are not able to measure party identification, since there is no way to know the individual preferences of voters. Nevertheless, based on how the result in a certain county turned out, we can get an idea about the actual preference of the majority of voters of that county.

Bartels [@Bartels2005] argues that "Democratic identification declined by 18% among low-income whites (from 22% in 1952 to 4% in 2004) and by 29% among high-income whites (from 11% to ???18%)." But he mentions that "However, it seems odd to attribute the Democrats' problems to the white working class when the corresponding decline among more affluent whites is so much larger." 

Although one individual has one vote, the value of that vote may be different. 
_can we see if the 18 percent decline among low-income whites (and seeing where they are concentrated) has more impact on the electoral outcome than the 29 percent decline among high-income whites who may be concentrated in bigger cities and urban areas and democratic states._





#Methodology:
##Dependent Variable:
Vote share (Republican Party) or Vote share (Democratic/Incumbent Party) or Difference in Voteshare between the two parties. In old literature, the second one is used often. However, in the modern literature the third one is used. In the case of this paper, the main question circles around Trump and hence the first variable may work best. Ask for advice?

##Independent Variable:
-Education (replace this with the data from census API)
-Demographic Data from Census (male, female, ethnicity, religion, education)
-Unemployment
-Per capita income
-Employment in industry
-Current Transfer Receipts
-Adjustment for Residence
-Dummy for incumbent party
-Rural or urban county

#Bibliograhy: